---
title: "Experimentally assigned condition discrimination"
description: "A governed demonstration using an experimentally assigned label."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimentally assigned condition discrimination}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gp3ml)
```


## Declared task

The label is the experimentally assigned condition. The workflow assesses whether predeclared measurements discriminate that assignment. It does not establish psychological interpretation or causal mechanism.

```{r task}
data <- simulate_gazepoint_governed_data(18L, 6L, 1L, seed = 2201L)
predictors <- c("fixation_duration", "gaze_dispersion", "pupil_change")
task <- create_gazepoint_synthetic_task(
  data, "assigned_condition", "new_participants"
)
manifest <- create_gazepoint_synthetic_manifest(task$outcome, predictors)
folds <- create_gazepoint_group_folds(
  data, task$outcome, predictors, manifest,
  task$generalization_target,
  task$participant_id, task$unit_id, task$stimulus_id,
  v = 3L, repeats = 1L, seed = 2201L
)
```

## Explicit candidate grid

```{r tune}
grid <- create_gazepoint_tuning_grid(
  engine = "glm",
  preprocessor_grid = list(center = c(TRUE, FALSE), scale = TRUE),
  thresholds = c(0.45, 0.55),
  complexity = "low",
  interpretability = "high"
)
tuning <- tune_gazepoint_model(
  folds, task, grid, predictors = predictors, seed = 2201L
)
compare_gazepoint_models(tuning, c("roc_auc", "balanced_accuracy", "brier"))
```

No candidate is selected automatically. A selection requires an explicit metric, direction, and human rationale.
